Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-09-21 manuscript/paper.tex
discussion and outlook. appendix: multipoles.

\tilde V_{lm} &= \sum_{l',m'}\mathcal{R}_{lm,l'm'}V_{l'm'},

@@

$\mathcal{R}(\alpha,\beta)$ is the rotation matrix for real spherical harmonics that we use (see appendix~\ref{sec:appendix}). 

@@

\tilde V_{lm} = \tilde V_{lm}^O + \sum_{l',m'}\mathcal{R}_{lm,l'm'} V_{l'm'}^{AB},\\
V_{l'm'}^{AB} = \left(\frac{q_A}{a} \right)V_{l'm'}^{A}   
+ \left(\frac{q_B}{a}\right)V_{l'm'}^{B},

@@

@@

Since $e_{g}$ has only two states,
ignoring the coupling between $e_{g}$ and $t_{2g}$,
we can always analytically diagonalise it.
The expressions for the
matrix elements of the Hamiltonian are long however,
so we present the expansion at small angles in the following.
The diagonal terms with reference to the average $e_g$ energy are given by
\begin{align}
\label{eq:eg-hiix}
(H_{x^2-y^2},&H_{z^2}) = (1,-1)
\left[\frac{}{} \Omega_{2}  D_2 + \Omega_4 D_4\right],\\
\Omega_{l} =&  \left(q_A \Omega_{l}^A + q_B\Omega_{l}^B\right)/a,~(l=2,4),\\
\Omega_{2}^A =&  -\frac{64 \left(\alpha ^2 \left(7 \beta ^2+3\right)-6 \beta ^2\right)}{189 \sqrt{3}},\\
\Omega_{4}^A =& \frac{640 \left(\alpha ^2 \left(\beta ^2+16\right)-32 \beta ^2\right)}{1701 \sqrt{3}},\\
\Omega_{2}^B =&  \frac{3}{14} \left(\alpha ^2 \left(3 \beta ^2+2\right)-4 \beta ^2\right),\\
\Omega_{4}^B =& -\frac{5}{168}\left(\alpha ^2 \left(37 \beta ^2+46\right)-92 \beta ^2\right),
\end{align}
whereas the coupling between the two states is,
\begin{align}
\label{eq:eg-hijx}
H_{x^2-y^2,z^2}&=
\frac{2\alpha ^2 \beta}{567} 
\left(128 q_A (3 D_2-20 D_4) - 81 \sqrt{3} q_B (2 D_2 -5 D_4)\right),
%&=\frac{2\alpha ^2 \beta}{567} \left(384 ( D_2-\frac{20}{3} D_4)q_A  - 162 \sqrt{3} (D_2 -\frac{5}{2} D_4)q_B \right)
\end{align}
which vanishes if 
either of $\alpha,\beta$ is zero.

@@

% tilt vs rotation, preference
In a given material,
assuming
the charges on A and B are fixed
and either A or B dominates,
one type of distortion 
would be energetically preferred
over the other
for certain
numbers of electrons in 
the $d$-orbitals of the Jahn-Teller ion B.
For example,
if B cage dominates,
a $d^2$ configuration
should prefer tilting over rotation
because it stabilizes two occupied orbitals 
while rotation stabilizes one and destabilizes the other occupied orbital.

% orbital ordering
We have focused on the B ion at $(0,0,0)$ but 
there are three more B ions in the unit cell 
at finite octahedral rotations.
We find that the splittings in the spectra
of all B ions are
practically identical.
However, their eigenstates 
are different in a way that 
creates interesting orbital ordering
which depends on the tilt and rotation angles,
unlike what happens in 
tetragonal Jahn-Teller distortions (elongation and compression) 
such as seen in KCuF$_3$, 
where the orbital ordering does not depend on the size of the distortion.
A detailed study of orbital ordering in this case is out of scope of this manuscript and is left for future work.

Of course, beside the crystal field we considered in our simplistic model,
in real materials,
 there would be other local effects like
electronic correlations (Hubbard and Hund's interaction), 
and spin-orbit interaction, 
as well as band structure effects arising from
the electron hopping between the B ions.
The interplay between these and the crystal field (or JT) effect 
we discussed in this manuscript is also an interesting future work.

Based on our analysis of A and B cages,
we expect that,
in case of atomic scale perovskite superlattices
with cations having different charges in alternating layers,
the crystal field at B sites will bear the exact same deformation as in case of 
a tetragonal distortion of the octahedra,
and should lead to 
a JT like splitting even 
in the symmetric structure, i.e., with no structural distortion.
Considering the charge as a dynamic variable,
it should then also be possible that in 
simple perovskites structure,
the system might encounter charge instabilities similar to the structural ones (i.e., JT),
where the system is stabilised by a charge transfer from one site to another
that lowers the electronic energy and costs 
the ``elastic'' energy proportional the chemical hardness of the ions involved.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\section{Appendix: Rotation matrices}
\label{sec:append}

\begin{gather}
\nonumber
R=
\left(
\scalemath{0.9}{
\begin{array}{ccc}
 \frac{1}{2} ((\cos \alpha+1) \cos \beta+(\cos \alpha-1) \sin \beta) & \frac{1}{2} (\cos \beta-\sin \beta-\cos \alpha (\cos \beta+\sin \beta)) & \frac{\sin \alpha (\cos \beta+\sin \beta)}{\sqrt{2}} \\
 \frac{1}{2} (-\cos \alpha \cos \beta+\cos \beta+\cos \alpha \sin \beta+\sin \beta) & \frac{1}{2} (\cos \beta+\cos \alpha (\cos \beta-\sin \beta)+\sin \beta) & \frac{\sin \alpha (\sin \beta-\cos \beta)}{\sqrt{2}} \\
 -\frac{\sin \alpha}{\sqrt{2}} & \frac{\sin \alpha}{\sqrt{2}} & \cos \alpha \\
\end{array}
}
\right),\\ \nonumber
\mathcal{R}=
\left(
\scalemath{0.65}{
\begin{array}{ccccc}
 \frac{1}{4} (\cos 2 \alpha+3) \cos 2 \beta & \frac{\sin \alpha (\cos \alpha \cos 2 \beta+\sin 2 \beta)}{\sqrt{2}} & -\frac{1}{2} \sqrt{3} \cos 2 \beta \sin ^2\alpha & \frac{\sin \alpha (\sin 2 \beta-\cos \alpha \cos 2 \beta)}{\sqrt{2}} & \cos \alpha \sin 2 \beta \\
 \frac{\cos \alpha \sin \alpha (\sin \beta-\cos \beta)}{\sqrt{2}} & \frac{1}{2} (\cos 2 \alpha (\cos \beta-\sin \beta)+\cos \alpha (\cos \beta+\sin \beta)) & \sqrt{\frac{3}{2}} \cos \alpha \sin \alpha (\sin \beta-\cos \beta) & \frac{1}{2} (\cos 2 \alpha (\sin \beta-\cos \beta)+\cos \alpha (\cos \beta+\sin \beta)) & -\frac{\sin \alpha (\cos \beta+\sin \beta)}{\sqrt{2}} \\
 -\frac{1}{2} \sqrt{3} \sin ^2\alpha & \sqrt{\frac{3}{2}} \cos \alpha \sin \alpha & \frac{1}{4} (3 \cos 2 \alpha+1) & -\sqrt{\frac{3}{2}} \cos \alpha \sin \alpha & 0 \\
 \frac{\cos \alpha \sin \alpha (\cos \beta+\sin \beta)}{\sqrt{2}} & \frac{1}{2} (\cos \alpha (\cos \beta-\sin \beta)-\cos 2 \alpha (\cos \beta+\sin \beta)) & \sqrt{\frac{3}{2}} \cos \alpha \sin \alpha (\cos \beta+\sin \beta) & \frac{1}{2} (\cos \alpha (\cos \beta-\sin \beta)+\cos 2 \alpha (\cos \beta+\sin \beta)) & \frac{\sin \alpha (\sin \beta-\cos \beta)}{\sqrt{2}} \\
 -\left(\cos ^2\alpha+1\right) \cos \beta \sin \beta & \frac{\sin \alpha (\cos 2 \beta-2 \cos \alpha \cos \beta \sin \beta)}{\sqrt{2}} & \sqrt{3} \cos \beta \sin ^2\alpha \sin \beta & \frac{\sin \alpha (\cos 2 \beta+\cos \alpha \sin 2 \beta)}{\sqrt{2}} & \cos \alpha \cos 2 \beta \\
\end{array}
}
\right)
\end{gather}

\section{Appendix}
\label{append:multipoles}
\sm{
It is worth noting that $V_{l,m}^{A/B}$ can be found without 
evaluating the 
inner products between the spherical harmonics and
$V^{A/B}(\vec r;\alpha,\beta)$.
The latter can be computed 
using the simple textbook formula 
$V^{A/B}(\vec r;\alpha,\beta)=\sum_{i}1/r_i$, $\vec r_i$ is the position of the observation point $\vec r$ (measured from the JT ion/central B ion).
Once we have ${V^{A/B}(\vec r;\alpha,\beta)}$,
$V_{l,m}^{A/B}$ can be calculated by first expanding this potential
in powers of $r$, 
and then expanding the coefficient of $r^l$ 
in powers of $\cos\theta, \cos\phi$ (or $\sin\theta, \sin\phi$; or simply the spherical coordinates $\theta,\phi$ of $\hat r$) 
and comparing the terms with similar 
expansion of the formal expression on the right side of Eq.~\ref{eq:ylm_expansion} to get linear equations in $\{V_{l,m}\}$, and finally solving this system of equations.
The only relevant non-zero components of $V_{l,m}^{A/B}$ are the following.

It is worth noting that $V_{l,m}^{A/B}$ can be found without 
evaluating the 
inner products between the spherical harmonics and
$V^{A/B}(\vec r;\alpha,\beta)$.
The latter can be computed 
using the simple textbook formula 
$V^{A/B}(\vec r;\alpha,\beta)=\sum_{i}1/r_i$, $\vec r_i$ is the position of the observation point $\vec r$ (measured from the JT ion/central B ion).
Once we have ${V^{A/B}(\vec r;\alpha,\beta)}$,
$V_{l,m}^{A/B}$ can be calculated by first expanding this potential
in powers of $r$, 
and then expanding the coefficient of $r^l$ 
in powers of $\cos\theta, \cos\phi$ (or $\sin\theta, \sin\phi$; or simply the spherical coordinates $\theta,\phi$ of $\hat r$) 
and comparing the terms with similar 
expansion of the formal expression on the right side of Eq.~\ref{eq:ylm_expansion}.
The only relevant non-zero components of $V_{l,m}^{A/B}$ are the following.

\subsection{$beta=0$ case: analytical solution:}

$E_{xy},E_{yz},g,\lambda$ all have the same form as 
\begin{align}
X = -\frac{1}{a}\sum_{l=2,4}\left(q_A X^{A,l} + q_B X^{B,l}\right) D_l,
\end{align}
%where $X=E_{xy},E_{yz},g,\lambda$,
where various terms are given in the following.

%\begin{align}
%E_{xy} = -\frac{1}{a}\sum_{l=2,4}\left(q_A E_{xy}^{A,l} + q_B E_{xy}^{B,l}\right) D_l,
%\end{align}
%where $E_{xy}^{A/B,l}$ are given in the following.

%A D2
\begin{align}
E_{xy}^{A,2} =& \frac{4 \left(-72 \cos 2 \alpha -27 \cos 4 \alpha +\sqrt{6} \sqrt{\cos 2 \alpha +5} \left(15 \cos \alpha +\cos 3 \alpha -16 \sec ^3\alpha \right)+99\right)}{63 \left(\cos ^2\alpha +2\right)^{5/2}},\\
E_{yz}^{A,2} =&\frac{144 \cos 2 \alpha +54 \cos 4 \alpha +\sqrt{\frac{3}{2}} (9 \sin \alpha +\sin 3 \alpha )^2 \sqrt{\cos 2 \alpha +5} \sec ^3\alpha -198}{63 \left(\cos ^2\alpha +2\right)^{5/2}},\\
g^{A,2} =&\frac{2 \left(2 \sqrt{3} (\cos 2 \alpha +5)^{5/2}-216 \sqrt{2} \cos ^5\alpha \right) \tan \alpha  \sec \alpha }{63 \left(\cos ^2\alpha +2\right)^{5/2}},\\
\lambda^{A,2} =&\frac{216 \cos 4 \alpha -4 \sqrt{6} (\cos 2 \alpha +3) (\cos 2 \alpha +5)^{5/2} \sec ^3\alpha +3240}{252 \left(\cos ^2\alpha +2\right)^{5/2}},
\end{align}

% A D4
\begin{align}
E_{xy}^{A,4} =& \frac{1}{20412 \left(\cos ^2\alpha +2\right)^{9/2}}
\left[\frac{}{} -330480 \cos 2 \alpha +243 (2964 \cos 4 \alpha +80 \cos 6 \alpha +35 \cos 8 \alpha +5449)\right.\\
&\hspace{4cm} + \left. \sqrt{\frac{3}{2}} (180 \cos 2 \alpha -245 \cos 4 \alpha +513) (\cos 2 \alpha +5)^{9/2} \sec ^5\alpha \right],\\
E_{yz}^{A,4} =& \frac{}{10206 \left(\cos ^2\alpha +2\right)^{9/2}}
\left[660960 \cos 2 \alpha +243 (652 \cos 4 \alpha -160 \cos 6 \alpha -35 \cos 8 \alpha +407)\right.\\
&\hspace{4cm} + \left.\sqrt{\frac{3}{2}} (60 \cos 2 \alpha +245 \cos 4 \alpha -81) (\cos 2 \alpha +5)^{9/2} \sec ^5\alpha \right],\\
g^{A,4} =& \frac{5 \left(31104 \sqrt{2} \cos ^7\alpha  (8 \cos 2 \alpha +7 \cos 4 \alpha +145)-32 \sqrt{3} (\cos 2 \alpha +5)^{9/2} (49 \cos 2 \alpha -9)\right) \tan \alpha  \sec ^3\alpha }{163296 \left(\cos ^2\alpha +2\right)^{9/2}},\\
\lambda^{A,4} =& \frac{1}{10206 \left(\cos ^2\alpha +2\right)^{9/2}}
\left[5 \left(180792 \cos 2 \alpha +243 (212 \cos 4 \alpha +24 \cos 6 \alpha +7 \cos 8 \alpha +37)\right)\right.\\
&\hspace{4cm} + \left.\sqrt{\frac{3}{2}} (12 \cos 2 \alpha -49 \cos 4 \alpha -27) (\cos 2 \alpha +5)^{9/2} \sec ^5\alpha \right],
\end{align}

% B D2
\begin{align}
E_{xy}^{B,2} =& \frac{32 \cos 2 \alpha +24 \cos 4 \alpha -(3 \cos 2 \alpha +1) (\cos 2 \alpha +3)^{5/2} \sec ^3\alpha +72}{7 (\cos 2 \alpha +3)^{5/2}},\\
E_{yz}^{B,2} =&\frac{-32 \cos 2 \alpha -24 (\cos 4 \alpha +3)+(3 \cos 2 \alpha +1) (\cos 2 \alpha +3)^{5/2} \sec ^3\alpha }{14 (\cos 2 \alpha +3)^{5/2}},\\
g^{B,2} =&\frac{3 \left(40 \cos \alpha +20 \cos 3 \alpha -\sqrt{\cos 2 \alpha +3} (12 \cos 2 \alpha +\cos 4 \alpha )+4 \cos 5 \alpha -19 \sqrt{\cos 2 \alpha +3}\right) \tan \alpha  \sec \alpha }{7 \sqrt{2} (\cos 2 \alpha +3)^{5/2}},\\
\lambda^{B,2} =&-\frac{3 \left(8 \cos 2 \alpha +(\cos 2 \alpha +3)^{3/2} \tan ^2\alpha  \sec \alpha -8\right)}{7 (\cos 2 \alpha +3)^{3/2}},
\end{align}

%B D4
\begin{align}
E_{xy}^{B,4} &= \scalemath{0.6}{\frac{1}{42} \left(-\frac{4 (851 \cos 2 \alpha +10 \cos 4 \alpha -35 \cos 6 \alpha +198)}{(\cos 2 \alpha +3)^{7/2}}+19 \sec ^5\alpha -50 \sec ^3\alpha +35 \sec \alpha \right)},\\
E_{yz}^{B,4} &= \scalemath{0.6}{\frac{-3648 \cos 2 \alpha -8 (628 \cos 4 \alpha +120 \cos 6 \alpha +35 \cos 8 \alpha +297)-\frac{1}{2} (20 \cos 2 \alpha +35 \cos 4 \alpha +9) (\cos 2 \alpha +3)^{9/2} \sec ^5\alpha }{84 (\cos 2 \alpha +3)^{9/2}}},\\
g^{B,4} &= \scalemath{0.6}{\frac{5 (5 \sin \alpha +\sin 3 \alpha ) \left(-3024 \cos \alpha -1344 \cos 3 \alpha +216 \cos 7 \alpha +56 \cos (9 \alpha )+(1224 \cos 2 \alpha +676 \cos 4 \alpha +120 \cos 6 \alpha +7 \cos 8 \alpha ) \sqrt{\cos 2 \alpha +3}+21 \sqrt{\cos 2 \alpha +3}\right) \sec ^4\alpha }{672 \sqrt{2} (\cos 2 \alpha +3)^{9/2}}},\\
\lambda^{B,4} &= \scalemath{0.6}{\frac{5}{42} \left(-\frac{32 \sin ^2\alpha  (4 \cos 2 \alpha +7 \cos 4 \alpha +5)}{(\cos 2 \alpha +3)^{7/2}}-(7 \cos 2 \alpha +5) \tan ^2\alpha  \sec ^3\alpha \right).}
\end{align}

\subsection{general case, t2g}

% using ham-approx-expansion-alpha-beta.nb: Generalt2gSplitAver[...]
\begin{align} 
\tilde H_{A,2} =& 
\left(
\begin{array}{ccc}
 -\frac{64 \left(\alpha ^2 \left(7 \beta ^2+3\right)-6 \beta ^2\right)}{189 \sqrt{3}} & \frac{32}{21} \sqrt{\frac{2}{3}} \alpha  \beta ^2 & -\frac{32}{21} \sqrt{\frac{2}{3}} \alpha  \beta ^2 \\
 & \frac{32 \left(\alpha ^2 (\beta  (7 \beta -12)+3)-6 \beta ^2\right)}{189 \sqrt{3}} & \frac{64 \alpha ^2 \left(3-2 \beta ^2\right)}{189 \sqrt{3}} \\
&  & \frac{32 \left(\alpha ^2 (\beta  (7 \beta +12)+3)-6 \beta ^2\right)}{189 \sqrt{3}} \\
\end{array}
\right),\\
\tilde H_{A,4} =& 
\left(
\begin{array}{ccc}
 -\frac{2560 \left(\alpha ^2 \left(\beta ^2+16\right)-32 \beta ^2\right)}{5103 \sqrt{3}} & \frac{1280 \sqrt{\frac{2}{3}} \alpha  \beta  (16 \beta -7)}{1701} & -\frac{1280 \sqrt{\frac{2}{3}} \alpha  \beta  (16 \beta +7)}{1701} \\
 & \frac{1280 \left(\alpha ^2 ((\beta -72) \beta +16)-32 \beta ^2\right)}{5103 \sqrt{3}} & -\frac{2560 \alpha ^2 \left(24 \beta ^2+1\right)}{1701 \sqrt{3}} \\
&& \frac{1280 \left(\alpha ^2 (\beta  (\beta +72)+16)-32 \beta ^2\right)}{5103 \sqrt{3}} \\
\end{array}
\right),\\
& 0,\frac{512 \left(5 \alpha ^2 \left(4 \beta ^2+1\right)+5 \beta ^2-3\right)}{729 \sqrt{3}}
\end{align}

\begin{align} 
\tilde H_{B,2} =& 
\left(
\begin{array}{ccc}
 \frac{3}{14} \left(\alpha ^2 \left(3 \beta ^2+2\right)-4 \beta ^2\right) & -\frac{9 \alpha  \beta ^2}{7 \sqrt{2}} & \frac{9 \alpha  \beta ^2}{7 \sqrt{2}} \\
 & -\frac{3}{28} \left(\alpha ^2 (\beta  (3 \beta -8)+2)-4 \beta ^2\right) & -\frac{3}{7} \alpha ^2 \left(\beta ^2+1\right) \\
  &  & -\frac{3}{28}\left(\alpha ^2 (\beta  (3 \beta +8)+2)-4 \beta ^2\right) \\
\end{array}
\right),\\
\tilde H_{B,4} =& 
\left(
\begin{array}{ccc}
 \frac{5}{126} \left(\alpha ^2 \left(37 \beta ^2+46\right)-92 \beta ^2\right) & \frac{5 \alpha  (14-23 \beta ) \beta }{21 \sqrt{2}} & \frac{5 \alpha  \beta  (23 \beta +14)}{21 \sqrt{2}} \\
 & -\frac{5}{252}\left(\alpha ^2 (\beta  (37 \beta -144)+46)-92 \beta ^2\right) & \frac{5}{21} \alpha ^2 \left(31 \beta ^2+8\right) \\
 & & -\frac{5}{252}\left(\alpha ^2 (\beta  (37 \beta +144)+46)-92 \beta ^2\right) \\
\end{array}
\right),\\
&0,\frac{1}{18} \left(-5 \alpha ^2 \left(7 \beta ^2+4\right)-20 \beta ^2+12\right) 
\end{align}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.48\textwidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/code-rigid-rotation/figs/theta-0-q-1-D24terms}
\includegraphics[width=0.48\textwidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/code-rigid-rotation/figs/theta-0-q-2-D24terms}
\includegraphics[width=0.48\textwidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/code-rigid-rotation/figs/theta-0-q-3-D24terms}
\caption{append fig; combine in one plot: $D_2/D_4\simeq 14$ at 
$a=7$ bohr and $2eV$ crystal field gap ($\braket{r^4}=11.5809$, $D_4\simeq0.005$, $D_2=\sqrt{D_4}\simeq0.07$.)
}
\label{default}
\end{center}
\end{figure}

\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{System and Model}
System: perovskites ABO3;
Full model = lattice;
Toy model = cages;

\section{Analytical Results}
phi rotations; 2x2 solutions;\\
theta/phi rotations; approx 2x2 
\subsection{Violation of Jahn-Teller Theorem}
\section{Numerical Results}
Exact numerics cages [cubic];;; exact lattice [with a,b,c lattice constants]; 
\section{Effect of SOC}

\begin{figure}
\centering
%\includegraphics[width=1.0\linewidth]{cartoon}
\caption{
}
\label{fig:cartoon}
\end{figure}

 %\end{document}
\label{Bibliography}
\bibliographystyle{apsrev4-1}
%\bibliography{References}

\com{
$V_{oct}(\vec r)$ causes the d-orbitals to split into two manifolds,
$t_{2g}$ ($xy,yz,zx$) and $e_g$ ($z^2,x^2-y^2$)
by an amount
${\Delta_{CF}=\frac{160}{3}\frac{q_o}{a} D_{4}}$.
The deformation and resulting $V_{dipoles}(\vec r)$
will lift the degeneracies as follows.
\begin{align}
(dE_{xy},dE_{yz},dE_{zx})&= \Delta_{\delta}
\left[(2, -1, -1)D_{2} +\frac{20}{9} 
 (-1, 4, 4) D_{4}\right],\\
(dE_{x^2-y^2},dE_{z^2})&=\Delta_{\delta}
\left[(2,-2)D_{2} +\frac{20}{9} 
 (-1,-6)D_{4}\right],\\
\Delta_{\delta}&= \frac{96}{7}
\frac{-q_o}{a}
\frac{\delta}{a}.
\end{align}
}

\com{
\begin{align}
\tilde V(\vec r) &=\tilde V_O(\vec r) + \tilde V_{AB}(\vec r),\\
\tilde V_O(\vec r) &= \tilde V_{oct}(\vec r),\\
\tilde V_{AB}(\vec r) &= \mathcal{R}(\alpha,\beta) V_{AB}(\vec r),\\
V_{AB}(\vec r) &= f(q_A) V_{oct}(\vec r) + 
V_{dipoles}^{AB}(\vec r;\alpha,\beta),\\
\label{eq:ylm_expansion}
V_{dipoles}^{AB}(\vec r;\alpha,\beta) &= 
\sum_{l=2,4}\sum_{m=-l}^{l}\left[V_{l,m}^{A}(\alpha,\beta)+V_{l,m}^{B}(\alpha,\beta)\right] r^l Y_{l,m}(\hat r),\\
\tilde V(\vec r) &= \sum_{l,m} \tilde V_{l,m} r^l \tilde Y_{l,m}(\hat r),\\\end{align}
\az{fix inconsistencies... in above equations.. R matix apply on a vector,,,, not a function.}.
}

>>>>

%\tilde V_{lm} &= \sum_{l',m'}\mathcal{R}_{lm,l'm'}V_{l'm'},
\tilde V_{lm} &= \sum_{m'=-l}^{l}\mathcal{R}^{l}_{m,m'}V_{lm'},

@@

$\mathcal{R}^{l}=\mathcal{R}^{l}(\alpha,\beta)$ is the rotation matrix for real spherical harmonics 
of degree $l$ (see appendix~\ref{sec:appendix}).

@@

\tilde V_{lm} = \tilde V_{lm}^O + \sum_{m'=-l}^{l}\mathcal{R}^{l}_{m,m'} V_{lm'}^{AB},\\
V_{lm}^{AB} = \left(\frac{q_A}{a} \right)V_{lm}^{A}   
+ \left(\frac{q_B}{a}\right)V_{lm}^{B},

@@

\az{

Summarise t2g results here.\\

copy from summary section...??

tilt vs rotation, preference:\\

In a given material,
assuming
the charges on A and B are fixed
and either A or B dominates,
one type of distortion 
would be energetically preferred
over the other
for certain
numbers of electrons in 
the $d$-orbitals of the Jahn-Teller ion B.
For example,
if B cage dominates,
a $d^2$ configuration
should prefer tilting over rotation
because it stabilises two occupied orbitals 
while rotation stabilises one and destabilises the other occupied orbital.

@@

Since $e_{g}$ has only two states,
ignoring the coupling between $e_{g}$ and $t_{2g}$,
we can always analytically diagonalise it.
The expressions for the
matrix elements of the Hamiltonian are long however,
so we present the expansion at small angles in the following.
The diagonal terms with reference to the average $e_g$ energy are given by
\begin{align}
\label{eq:eg-hiix}
(H_{x^2-y^2},&H_{z^2}) = (1,-1)
\left[\frac{}{} \Omega_{2}  D_2 + \Omega_4 D_4\right],\\
\Omega_{l} =&  \left(q_A \Omega_{l}^A + q_B\Omega_{l}^B\right)/a,~(l=2,4),\\
\Omega_{2}^A =&  -\frac{64 \left(\alpha ^2 \left(7 \beta ^2+3\right)-6 \beta ^2\right)}{189 \sqrt{3}},\\
\Omega_{4}^A =& \frac{640 \left(\alpha ^2 \left(\beta ^2+16\right)-32 \beta ^2\right)}{1701 \sqrt{3}},\\
\Omega_{2}^B =&  \frac{3}{14} \left(\alpha ^2 \left(3 \beta ^2+2\right)-4 \beta ^2\right),\\
\Omega_{4}^B =& -\frac{5}{168}\left(\alpha ^2 \left(37 \beta ^2+46\right)-92 \beta ^2\right),
\end{align}
whereas the coupling between the two states is,
\begin{align}
\nonumber
H_{x^2-y^2,z^2}=
\frac{2\alpha ^2 \beta}{567} 
&\left(\frac{}{}128 q_A (3 D_2-20 D_4) \right. \\ \label{eq:eg-hijx}
&~~\left. - 81 \sqrt{3} q_B (2 D_2 -5 D_4)\right),
%&=\frac{2\alpha ^2 \beta}{567} \left(384 ( D_2-\frac{20}{3} D_4)q_A  - 162 \sqrt{3} (D_2 -\frac{5}{2} D_4)q_B \right)
\end{align}
which vanishes if 
either of $\alpha,\beta$ is zero.

@@

\com{
% beside the hardness, the Madelung energy will increase due to an increased repulsion between the two ions that take part in the charge transfer. Consider that q1^2+q2^2 with constant that q1+q2=2q=fixed is min when q1=q2=q.
% so the electronic energy gain will have to fight against both the chem hardness and the Madelung energy.
Based on our analysis of A and B cages,
we expect that,
in case of atomic scale perovskite superlattices
with cations having different charges in alternating layers,
the crystal field at B sites will bear the exact same deformation as in case of 
a tetragonal distortion of the octahedra,
and should lead to 
a JT like splitting even 
in the symmetric structure, i.e., with no structural distortion.
Considering the charge as a dynamic variable,
it should then also be possible that in 
simple perovskites structure,
the system might encounter charge instabilities similar to the structural ones,
where the system is stabilised by a charge transfer from one site to another
that lowers the electronic energy and costs 
the ``elastic'' energy proportional to the chemical hardness of the ions involved.
Such considerations might be useful in studying polaron formation and dynamics.
}

% orbital ordering
We have only focused on the spectrum of the B ion at 
%position ${(0,0,0)}$
origin
 in the cell 
described in sec.~\ref{sec:structure}
but 
there are three more \emph{distinct} B ions in the unit cell 
at finite octahedral rotations that also 
need to be considered simultaneously.
We find that the splittings in the spectra
of all B ions are
practically identical.
However, their eigenstates 
are different in a way that can
create interesting orbital ordering
that depends on the tilt and rotation angles,
unlike what happens in 
tetragonal Jahn-Teller distortions (elongation and compression) 
such as seen in KCuF$_3$, 
where the orbital ordering does not depend on the size of the distortion.
The eigenstates' components in the local bases attached to the octahedra's frame
also vary between different B sites 
so the dependence on the rotation and tilt angle is not trivial,
which is understandable because different B sites in the unit cell have different environments. 
A detailed study of orbital ordering in this case is out of scope of this manuscript and is left for future work.

Of course, beside the crystal field we consider in our simplistic model,
in real materials,
 there would be other local effects like
electronic correlations (Hubbard and Hund's interaction), 
and spin-orbit interaction, 
as well as band structure effects arising from
the electron hopping between the B ions.
The interplay between these and the crystal field (or JT) effect 
we discussed in this manuscript is also an interesting future work.

%\begin{acknowledgements}
%\end{acknowledgements}

\appendix*
\section{Computing $V_{l,m}^{A/B}$}
\label{append:multipoles}

Here we describe the procedure to obtain the multipole components 
$V_{l,m}^{A/B}$ of potentials $V^{A/B}(\vec r)$ of $A/B$ ion cages 
at a position $\vec r$ around a JT ion.
That is, we like to expand the potentials $V^{A/B}(\vec r)$
in spherical harmonics as
\begin{align}
\label{eq:ylm_compA}
V^{A}(\vec r) =& \sum_{l,m} V_{lm}^{A} r^l Y_{lm}(\hat r),\\
\label{eq:ylm_compB}
V^{B}(\vec r) =& \sum_{l,m} V_{lm}^{B} r^l Y_{lm}(\hat r).
\end{align}
It is worth noting that $V_{l,m}^{A/B}$ can be found without 
evaluating the 
inner products between the spherical harmonics and
$V^{A/B}(\vec r)=V^{A/B}(\vec r;\alpha,\beta)$, as follows.

First we calculate
$V^{A/B}(\vec r)$
using the simple textbook formula.
$V^{A}(\vec r;\alpha,\beta)=\sum_{i}1/|\vec r^A_i-\vec r |$, 
where $\vec r^A_i$ is the position of $i$-th A ion
and 
$\vec r$ is the position of the observation point 
measured from the JT ion.
Similarly,
$V^{B}(\vec r;\alpha,\beta)=\sum_{i}1/|\vec r^B_i-\vec r |$.
Writing cartesian components of $\vec r$ as
$\vec r=r(\cos\phi\sin\theta,\sin\theta\sin\phi,\cos\theta)$,
where $(r,\theta,\phi)$ 
are the spherical coordinates in the frame of JT ion,
and expanding 
${V^{A/B}(\vec r;\alpha,\beta)}$
in powers of $r$, 
we can separate 
coefficient of $r^l$
that give degree $l$ components.
We can now consider each degree $l$ term individually,
and resolve it in its $(2l+1)$ $m$-components.
This can be done by 
expanding such a coefficient of $r^l$
in powers of $\cos\theta, \cos\phi$ 
(or $\sin\theta, \sin\phi$, or simply $\theta,\phi$) 
and comparing the terms with a similar expansion of 
the formal expression on the right side 
of Eq.~\ref{eq:ylm_compA} or Eq.~\ref{eq:ylm_compB}.
This gives a set of
linear equations in $\{V_{l,m}^{A/B}\}$ (for a fixed value of $l$)
 that can be easily solved.
The whole procedure can be performed
 on Mathematica~\cite{Mathematica}
or a similar software.

@@
